---
title: "RSCanner Script"
author: "Gandhar Mahadeshwar"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
RSCanner script [property of Pyle Lab, Yale University]
Written by Gandhar Mahadeshwar, Rafael Tavares

# Load the required packages
```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

## Step 1: Inputs
# Any of the inputs marked "USER" are either input files or parameters that can be freely adjusted by the user
```{r}
#USER: Input dot-bracket file
#dotbracket file must have three lines: title line, sequence line, dotbracket line, and no newline characters except 
#for the ones separating these three lines.

dotbracketinput <- readLines("sample_test_files/lncRNA-GRS1.dot")[3]
dotbracket_vector <- strsplit(dotbracketinput, split = NULL)[[1]]

#USER: Input ct file. Uncomment this to load in the ct file example instead.
#ctfileinput <- read.table("HCV_Jc1.ct", skip = 1)$V5

#USER (OPTIONAL): Input list of Shannon values
shannoninput <- read.table("sample_test_files/lncRNA-GRS1_shannon.txt", header = FALSE)
shannon <- shannoninput$V2
#USER: Input window size for BPC calculation
window_size <- 51
#USER: Input window size for Shannon entropy smoothing
window_size_shan <- 51
#USER: Input Shannon Entropy cutoff
SE_cutoff <- 0.5
#USER: Input BPC cutoff
BPC_cutoff <- 0.5
#USER: Input integer window length for histogram and heatmap
finalwind <- 100
#USER: Input the x-axis bounds (nt) for visualization
x_low_bound <- 1 # if no argument, default to 1
x_upper_bound <-length(shannon) # if no argument, default to length(shannon)
```

# Convert input ct file to dotbracket vector format. Run these lines only if the input structure file is a ct file, NOT a dotbracket file.
```{r}
dotbracket_nums <- ctfileinput

indexnotzero <- which(dotbracket_nums != 0)
for (i in 2:length(indexnotzero)) {
  if(indexnotzero[i] - indexnotzero[i-1] != 1) {
    dotbracket_nums[indexnotzero[i]] <- "*"
  }
  dotbracket_nums[indexnotzero[1]] <- "*"
}
wichaster <- which(dotbracket_nums == "*")
for (j in 1:length(wichaster)) {
  if (j %% 2 == 0) {
    dotbracket_nums[wichaster[j]] <- ")"
  } else {
    dotbracket_nums[wichaster[j]] <- "("
  }
}
dotbracket_real <- dotbracket_nums
`%!in%` <- Negate(`%in%`)
for (k in 1:length(dotbracket_nums)) {
  ind <- wichaster[which(wichaster < k)[length(which(wichaster < k))]]
  if (dotbracket_real[k] %!in% c("0", "(", ")")) {
    dotbracket_real[k] <- dotbracket_nums[ind]
  } else {
    dotbracket_real[k] <- dotbracket_nums[k]
  }
}
dotbracket_real[which(dotbracket_real == "0")] <- "."
dotbracket_vector <- dotbracket_real
```

## Step 2: Sliding window calculations
# Base pair content (BPC) calculation with customized sliding window size
```{r}
dotbracket_vector_bin <- numeric(length(dotbracket_vector))
for (i in 1:length(dotbracket_vector)) {
 if (dotbracket_vector[i] == ".") {
   dotbracket_vector_bin[i] <- 1
 } else {
   dotbracket_vector_bin[i] <- 0
 }
}
if ((window_size %% 2) == 0) {
  gapnum <- window_size/2 + 1
} else {
  gapnum <- (window_size+1)/2 
}
window_initvec <- numeric(length(dotbracket_vector))
window_termvec <- numeric(length(dotbracket_vector))
window_termvec[1] <- gapnum
for (j in 1:gapnum) {
  window_initvec[j] <- 1
}
for (k in (gapnum + 1):(length(dotbracket_vector))){
  window_initvec[k] <- window_initvec[k-1] + 1
}
for (l in (length(dotbracket_vector) - gapnum + 1):(length(dotbracket_vector))) {
  window_termvec[l] <- length(dotbracket_vector)
}
for (m in 2:(length(dotbracket_vector) - gapnum)) {
  window_termvec[m] <- window_termvec[m-1] + 1
}
dotcount <- numeric(length(dotbracket_vector))
for (o in 1:length(dotbracket_vector)) {
  dotcount[o] <- sum(dotbracket_vector_bin[window_initvec[o]:window_termvec[o]])
}
dotperc <- numeric(length(dotbracket_vector))
for (p in 1:length(dotbracket_vector)) {
  dotperc[p] <- (dotcount[p])/(window_termvec[p] - window_initvec[p] + 1)
}
oneminus_dotperc <- 1 - dotperc
```

# Plot BPC along the full length of the RNA
```{r}
bpcplotdata <- as.data.frame(cbind(seq(1, length(oneminus_dotperc)), oneminus_dotperc)) %>% rename("Nucleotide Position" = V1) %>% 
  rename("Base Pair Content" = oneminus_dotperc)
ggplot(data = bpcplotdata, aes(x = `Nucleotide Position`, y = `Base Pair Content`)) + theme_classic() + 
  geom_line() + 
  geom_hline(yintercept = quantile(oneminus_dotperc, probs=c(BPC_cutoff),name=FALSE), linetype = "dashed", color = "black", size = 1) + 
  scale_x_continuous(limits = c(x_low_bound, x_upper_bound))+
  theme(axis.text = element_text(size = 10, color="black"), axis.title = element_text(size = 12), panel.border = element_rect(color="black", fill=NA, size = 1))
```

# Save BPC data as a csv
```{r}
write.csv(bpcplotdata, "bpc_data.csv")
```

# Shannon entropy (SE) smoothing with customized sliding window size
```{r}
if ((window_size_shan %% 2) == 0) {
  gapnum_shan <- window_size_shan/2 + 1
} else {
  gapnum_shan <- (window_size_shan+1)/2 
}
window_initvec_shan <- numeric(length(shannon))
window_termvec_shan <- numeric(length(shannon))
window_termvec_shan[1] <- gapnum_shan
for (j in 1:gapnum_shan) {
  window_initvec_shan[j] <- 1
}
for (k in (gapnum_shan + 1):(length(shannon))){
  window_initvec_shan[k] <- window_initvec_shan[k-1] + 1
}
for (l in (length(shannon) - gapnum_shan + 1):(length(shannon))) {
  window_termvec_shan[l] <- length(shannon)
}
for (m in 2:(length(shannon) - gapnum_shan)) {
  window_termvec_shan[m] <- window_termvec_shan[m-1] + 1
}
med_shan <- numeric(length(shannon))
for (o in 1:length(shannon)) {
  med_shan[o] <- median(shannon[window_initvec_shan[o]:window_termvec_shan[o]])
}
```

# Plot smoothed SE along the full length of the RNA
```{r}
shanplotdata <- as.data.frame(cbind(seq(1, length(med_shan)), med_shan)) %>% rename("Nucleotide Position" = V1) %>% 
  rename("Smoothed Shannon Entropy" = med_shan)
ggplot(data = shanplotdata, aes(x = `Nucleotide Position`, y = `Smoothed Shannon Entropy`)) + theme_classic() +
  geom_line() + 
  geom_hline(yintercept = quantile(med_shan, probs=c(SE_cutoff),name=FALSE), linetype = "dashed", color = "black", size = 1) + 
  scale_x_continuous(limits = c(x_low_bound, x_upper_bound))+
  theme(axis.text = element_text(size = 10, color="black"), axis.title = element_text(size = 12), panel.border = element_rect(color="black", fill=NA, size = 1))
```

# Save smoothed SE data as a csv
```{r}
write.csv(shanplotdata, "smoothed_Shannon_data.csv")
```

## Step 3: Filtering
# Filter nucleotides that have BPC above the threshold and smoothed SE below the threshold ("structure counts")
```{r}
a <- which(oneminus_dotperc > quantile(oneminus_dotperc, probs=c(BPC_cutoff),name=FALSE)) #indices of the bpcs that are above cutoff
b <- which(med_shan < quantile(med_shan, probs=c(SE_cutoff),name=FALSE)) #indices of shannons that are below cutoff
abinter <- intersect(a,b) #intersection of the two vectors

passing_nucs <- as.data.frame(cbind(`Nucleotide` = abinter, 
                                    `Base Pair Content` = oneminus_dotperc[abinter], 
                                    `Shannon` = med_shan[abinter]))
```

# Save csv of nucleotides that pass both thresholds
```{r}
write.csv(passing_nucs, "structure_counts.csv")
```

# Output proportion of nucleotides that meet threshold
```{r}
length(abinter)/length(med_shan)
```

# Plot histogram of structure counts along the RNA
```{r}
ggplot(data=(new_dat %>% rename(`Structure Counts` = vals) %>% rename(`Nucleotide` = pos)), aes(x=`Nucleotide`, y=`Structure Counts`)) +
  geom_bar(stat="identity",  fill="grey", colour="black", width=100)+
  theme_classic()+
  theme(axis.text = element_text(size = 10, color="black"), axis.title = element_text(size = 12), panel.border = element_rect(color="black", fill=NA, size = 1))
```

## Step 4: Binning and Visualization
# Note to user: The bin size here is hard-coded to 100 nt. If you want to adjust this parameter, we recommend using the included R scripts.
```{r}
if ((finalwind %% 2) == 1) {
  finalwind <- finalwind + 1
} else {
  finalwind <- finalwind
}

finalwind_init <- numeric(ceiling(length(shannon)/finalwind))
finalwind_fin <- numeric(ceiling(length(shannon)/finalwind))
finalwind_init[1] <- 1
for (k in 2:length(finalwind_init)) {
  finalwind_init[k] <- finalwind_init[k-1] + finalwind
}
finalwind_fin[1] <- finalwind
for (k in 2:length(finalwind_fin)) {
  finalwind_fin[k] <- finalwind_fin[k-1] + finalwind
}
finalwind_fin[length(finalwind_fin)] <- length(shannon)
structure_counts <- numeric(length(finalwind_fin))
for (o in 1:length(structure_counts)) {
  count_raw <- sum(finalwind_init[o]<=abinter & abinter<=finalwind_fin[o])
  window_size <- (finalwind_fin[o] - finalwind_init[o] + 1)
  structure_counts[o] <- 100*(count_raw/window_size)
}
finalwind_inds <- seq(from = finalwind/2, to = length(shannon), by = finalwind)

if (length(finalwind_inds) != length(structure_counts)) {
  finalwind_inds_real <- numeric(length(finalwind_inds)+1)
  finalwind_inds_real[1:length(finalwind_inds_real)-1] <- finalwind_inds
  finalwind_inds_real[length(finalwind_inds_real)] <- finalwind_inds[length(finalwind_inds)]+finalwind
} else {
  finalwind_inds_real <- finalwind_inds
  }
```

# Output ordered table of structure content
```{r}
bin_number <- seq(from = 1, to = length(finalwind_inds_real), by = 1)
unordered_results_table <- as.data.frame(cbind(bin_number, structure_counts, finalwind_init, finalwind_fin))
ordered_results_table <- arrange(unordered_results_table, desc(structure_counts)) %>% rename("Bin Number" = bin_number) %>%
  rename("% Structure Content" = structure_counts) %>% rename("Bin Start (nt)" = finalwind_init) %>% 
  rename("Bin End (nt)" = finalwind_fin)

ordered_results_table

#Save the ordered table to a csv
write.csv(ordered_results_table, "ordered_structure_table.csv")
```

# Plot the percentage of structure content in non-overlapping bins along the RNA and overlay with a heatmap
```{r}
half_finalwind <- finalwind/2

#color ramp creation
colorramp <-  colorRampPalette(colors=c("#FFFF00", "#FF0000"))(finalwind+1)
inds_colors <- numeric(length(finalwind_inds_real))
for (i in 1:length(structure_counts)) {
  inds_colors[i] <- colorramp[structure_counts[i]+1]
}
inds_colors[which(inds_colors == "#FFFF00")] <- "#FFFFFF"
dat <- data.frame(pos = finalwind_inds_real, vals = structure_counts, cols = inds_colors)
starty <- finalwind_inds_real - half_finalwind
endy <- finalwind_inds_real + half_finalwind
endy[length(endy)] <- length(shannon)
## highlight region data
rects <- data.frame(start=starty, end=endy, group=seq_along(starty))

rects_lowwy <- which(rects$start == floor(x_low_bound/finalwind)*100)

if ((ceiling(x_upper_bound/finalwind)*100) > rects[nrow(rects),]$end) {
  rects_uppy <- which(rects$end == rects[nrow(rects),]$end)
} else {
  rects_uppy <- which(rects$end == ceiling(x_upper_bound/finalwind)*100)
}

new_rects <- rects %>% slice(rects_lowwy:rects_uppy)

new_dat <- dat %>% slice(rects_lowwy:rects_uppy)

#min and max x and y values
ymin <- min(dat$vals)
ymax <- max(dat$vals)
xmin <- new_rects$start[1]
xmax <- new_rects$end[length(new_rects$end)]

#generate line plot with heatmap overlay
heatmap <- ggplot(data=new_dat, aes(pos, vals)) +
  theme_classic() +
  geom_rect(data=new_rects, inherit.aes=FALSE, aes(xmin=starty[rects_lowwy:rects_uppy], xmax=endy[rects_lowwy:rects_uppy], ymin=ymin,
                                               ymax=ymax, group=group), color="transparent", 
            fill=as.character(new_dat$cols), alpha=1) + geom_line(lty=1, color="black", size = 0.9) +
            coord_cartesian(xlim = c(xmin, xmax+half_finalwind), ylim = c(ymin, ymax)) +
  xlab("Nucleotide") + ylab("% Structure Content") + theme(axis.text = element_text(size = 10, color="black"), axis.title = element_text(size = 12), panel.border = element_rect(color="black", fill=NA, size = 1))

#display plot
heatmap
```

# Run this block if you want to save the final line plot/heatmap to a high quality image
```{r}
## USER: Specify the parameters for the final plot
filename <- "heatmap"
device <- "tiff"
path <- NULL # defaults to current working directory
width <- 7
height <- 3
dpi <- 300
##
ggsave(filename, device=device, path=path, width=width, height=height, dpi=dpi)
```
