---
title: "RSCanner Script"
author: "Gandhar Mahadeshwar"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
RSCanner script [property of Pyle Lab, Yale University]
Written by Gandhar Mahadeshwar, Rafael Tavares

```{r}
library(readxl)
library(MASS)
library(tidyverse)
library(dplyr)
library(grid)
library(boxplotdbl)
library(xlsx)
library(seqinr)
```

#loading of the data
USER INPUTS: JFH1 HCV genomic RNA TEST-- uncomment this when you want to see JFH1 HCV genomic RNA results
```{r}

#USER: Input dot-bracket file
dotbracketinput <- seqinr::read.fasta("HCV_JFH1_2a.txt", set.attributes = FALSE, whole.header = FALSE)
dotbracket_vector <- dotbracketinput$ENERGY[((length(dotbracketinput$ENERGY)/2)+1):length(dotbracketinput$ENERGY)]

#USER: Input ct file
ctfileinput <- read.table("HCV_JFH1_2a_genome_model.ct")

#USER (OPTIONAL): Input list of Shannon values
shannoninput <- read.table("shannon.txt", header = TRUE)
shannon <- shannoninput$Shannon

#USER: Input window size for BPC calculation
window_size <- 50

#USER: Input window size for Shannon entropy smoothing
window_size_shan <- 50

```

#convert ct file to dot bracket vector, Run those lines if input structure file is a ct file
```{r}
dotbracket_nums <- ctfileinput[[4]]
#dotbracket_nums[which(dotbracket_nums == 0)] <- "."

# dotbracket_real <- numeric(length(dotbracket_nums))
# 
# 
# for (i in 1:length(dotbracket_real)) {
#   while (dotbracket_nums[i] != 0) {
#     dotbracket[i] <- brack_current
#   }
# }

indexnotzero <- which(dotbracket_nums != 0)

for (i in 2:length(indexnotzero)) {
  if(indexnotzero[i] - indexnotzero[i-1] != 1) {
    dotbracket_nums[indexnotzero[i]] <- "*"
  }
  dotbracket_nums[indexnotzero[1]] <- "*"
}

wichaster <- which(dotbracket_nums == "*")

for (j in 1:length(wichaster)) {
  if (j %% 2 == 0) {
    dotbracket_nums[wichaster[j]] <- ")"
  } else {
    dotbracket_nums[wichaster[j]] <- "("
  }
}

dotbracket_real <- dotbracket_nums

`%!in%` <- Negate(`%in%`)

for (k in 1:length(dotbracket_nums)) {
  ind <- wichaster[which(wichaster < k)[length(which(wichaster < k))]]
  if (dotbracket_real[k] %!in% c("0", "(", ")")) {
    dotbracket_real[k] <- dotbracket_nums[ind]
  } else {
    dotbracket_real[k] <- dotbracket_nums[k]
  }
}

dotbracket_real[which(dotbracket_real == "0")] <- "."

## USE ^^^ dotbracket_real for "dotbracket_vector" downstream script if ctfile is the input.
dotbracket_vector <- dotbracket_real
```

#base pair content calculation using customized window size
```{r}
dotbracket_vector_bin <- numeric(length(dotbracket_vector))
for (i in 1:length(dotbracket_vector)) {
 if (dotbracket_vector[i] == ".") {
   dotbracket_vector_bin[i] <- 1
 } else {
   dotbracket_vector_bin[i] <- 0
 }
}

if ((window_size %% 2) == 0) {
  gapnum <- window_size/2 + 1
} else {
  gapnum <- (window_size+1)/2 + 1
}

window_initvec <- numeric(length(dotbracket_vector))
window_termvec <- numeric(length(dotbracket_vector))
window_termvec[1] <- gapnum

for (j in 1:gapnum) {
  window_initvec[j] <- 1
}

for (k in (gapnum + 1):(length(dotbracket_vector))){
  window_initvec[k] <- window_initvec[k-1] + 1
}

for (l in (length(dotbracket_vector) - gapnum + 1):(length(dotbracket_vector))) {
  window_termvec[l] <- length(dotbracket_vector)
}

for (m in 2:(length(dotbracket_vector) - gapnum)) {
  window_termvec[m] <- window_termvec[m-1] + 1
}

dotcount <- numeric(length(dotbracket_vector))
for (o in 1:length(dotbracket_vector)) {
  dotcount[o] <- sum(dotbracket_vector_bin[window_initvec[o]:window_termvec[o]])
}

dotperc <- numeric(length(dotbracket_vector))
for (p in 1:length(dotbracket_vector)) {
  dotperc[p] <- (dotcount[p])/(window_termvec[p] - window_initvec[p] + 1)
}

oneminus_dotperc <- 1 - dotperc
```


```{r}
#Shannon smoothing with customized size sliding window



#movmedian calculation
if ((window_size_shan %% 2) == 0) {
  gapnum_shan <- window_size_shan/2 + 1
} else {
  gapnum_shan <- (window_size_shan+1)/2 + 1
}

window_initvec_shan <- numeric(length(shannon))
window_termvec_shan <- numeric(length(shannon))
window_termvec_shan[1] <- gapnum_shan

for (j in 1:gapnum_shan) {
  window_initvec_shan[j] <- 1
}

for (k in (gapnum_shan + 1):(length(shannon))){
  window_initvec_shan[k] <- window_initvec_shan[k-1] + 1
}

for (l in (length(shannon) - gapnum_shan + 1):(length(shannon))) {
  window_termvec_shan[l] <- length(shannon)
}

for (m in 2:(length(shannon) - gapnum_shan)) {
  window_termvec_shan[m] <- window_termvec_shan[m-1] + 1
}

med_shan <- numeric(length(shannon))
for (o in 1:length(shannon)) {
  med_shan[o] <- median(shannon[window_initvec_shan[o]:window_termvec_shan[o]])
}
```

```{r}
a <- which(oneminus_dotperc > median(oneminus_dotperc)) #indices of the bpcs that are above median
b <- which(med_shan < median(med_shan)) #indices of shannons that are below median

abinter <- intersect(a,b) #intersection of the two vectors
```
#output proportion of nucleotides that meet threshold (upstream sanity check for user)

```{r}
#example of intermediary output here:
length(abinter)/length(med_shan)
```


```{r}
plot(oneminus_dotperc, type = 'l')
```

```{r}
plot(med_shan, type = 'l')
```

```{r}
finalwind <- 100

finalwind_init <- numeric(ceiling(length(shannon)/finalwind))
finalwind_fin <- numeric(ceiling(length(shannon)/finalwind))

finalwind_init[1] <- 1
for (k in 2:length(finalwind_init)) {
  finalwind_init[k] <- finalwind_init[k-1] + 100
}

finalwind_fin[1] <- 100
for (k in 2:length(finalwind_fin)) {
  finalwind_fin[k] <- finalwind_fin[k-1] + 100
}

finalwind_fin[length(finalwind_fin)] <- length(shannon)

structure_counts <- numeric(length(finalwind_fin))
for (o in 1:length(structure_counts)) {
  count_raw <- sum(finalwind_init[o]<=abinter & abinter<=finalwind_fin[o])
  window_size <- (finalwind_fin[o] - finalwind_init[o] + 1)
  structure_counts[o] <- 100*(count_raw/window_size)
}

finalwind_inds <- seq(from = 50, to = length(shannon), by = 100)
```

```{r}
plot(x = finalwind_inds, y = structure_counts, type = 'l')
```

```{r}
which(structure_counts > 90)
which(structure_counts > 75 & structure_counts < 90)
which(structure_counts > 50 & structure_counts < 75)
```

```{r}
# data_unordered <- data.frame(pos = finalwind_inds, vals = structure_counts)
# dato <- arrange(data_unordered, desc(vals))
# dat_with_cols <- dato %>% mutate(cols = colorRampPalette(colors=c("#FF0000", "#FFFF00"))(length(finalwind_inds)))
# dat <- arrange(dat_with_cols, pos)
# 
# dat$cols[which(dat$vals == 0)] <- "#FFFFFF"
#FFFF00
#color ramp creation
colorramp <-  colorRampPalette(colors=c("#FFFF00", "#FF0000"))(101)
inds_colors <- numeric(length(finalwind_inds))

for (i in 1:length(structure_counts)) {
  inds_colors[i] <- colorramp[structure_counts[i]+1]
}

inds_colors[which(inds_colors == "#FFFF00")] <- "#FFFFFF"

dat <- data.frame(pos = finalwind_inds, vals = structure_counts, cols = inds_colors)

start <- finalwind_inds - 50
end <- finalwind_inds + 50
end[length(end)] <- length(shannon)

## highlight region data
rects <- data.frame(start=start, end=end, group=seq_along(start))

library(ggplot2)
d <- ggplot(data=dat, aes(pos, vals)) +
  theme_minimal() +
  geom_rect(data=rects, inherit.aes=FALSE, aes(xmin=start, xmax=end, ymin=min(dat$vals),
                ymax=max(dat$vals), group=group), color="transparent", 
            fill=as.character(dat$cols), alpha=1) + geom_line(lty=1, color="black", size = 0.9) +
  xlab("Nucleotide") + ylab("% High BPC/Low Shannon Entropy")
```


```{r}
#changing axis bounds to "zoom in" on region (1-2500)
#output text file of generated data - bpc, shannon, final file of well defined structure
#automate CT file conversion to dot bracket structure file
#static output with 10 discrete colors

d
```

```{r}
colorRampPalette(colors=c("#FF0000", "#FFFF00"))(length(dat$vals))
```


----- the next two blocks are representative tests showcasing some graph-building to be workshopped ----

```{r}
## Example data - this is for testing...
set.seed(0)
dat <- data.frame(dates=seq.Date(Sys.Date(), Sys.Date()+99, 1),
                  value=cumsum(rnorm(100)))

## Determine highlighted regions
v <- rep(0, 100)
v[c(5:20, 30:35, 90:100)] <- 1

## Get the start and end points for highlighted regions
inds <- diff(c(0, v))
start <- dat$dates[inds == 1]
end <- dat$dates[inds == -1]
if (length(start) > length(end)) end <- c(end, tail(dat$dates, 1))

## highlight region data
rects <- data.frame(start=start, end=end, group=seq_along(start), coloring=c(1.2, 3.2, 4.8))

library(ggplot2)
ggplot(data=dat, aes(dates, value)) +
  theme_minimal() +
  geom_line(lty=2, color="steelblue", lwd=1.1) +
  geom_point() +
  geom_rect(data=rects, inherit.aes=FALSE, aes(xmin=start, xmax=end, ymin=min(dat$value),
                ymax=max(dat$value), group=group), color="transparent", fill="orange", alpha=0.3) + geom_tile()
```

```{r}
x<-c(4.80,1.09,33.40,0.00,11.63,1.86,2.22,-10.95,-3.09,0.00,2.38,0.00,
         43.46,12.55,-49.51,0.00,0.00,0.00,1.72,0.00,0.00,-0.74,0.00,-5.79)

    image(matrix(x), col=colorRampPalette(colors=c("#FF0000", "#FFFF00"))(length(x)))
    image(matrix(x), col=colorRampPalette(colors=c("#FF0000", "#FFFF00"))(length(unique(x))))
```

